Investigations of thermal reactionomes reveals stress tolerance vs. resistance

### 2014-08-08 doc/ApTranscriptome_manuscript.Rmd
Updated with comments from NS

---
title: "Patterns of stress tolerance versus resistance in the thermal reactionomes of cold and warm-climate ant species"
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

**250 words allowed - need to cut 26**

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. In this study, we demonstrate how insight on the evolution of thermal tolerance can be gained through characterizing an organism’s thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient. We characterized the reactionomes of two populations of ant species of eastern U.S. forests: the northern cold-climate Aphaenogaster picea and the southern warm-climate Aphaenogaster carolinensis . We found that  at least 10% of genes were thermally responsive. When the two species were compared, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression,, upregulated of fewer heat shock proteins, and had greater inertia of changes in gene expression in response to thermal extremes. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. A surprisingly large proportion of genes showed divergent patterns of response to temperature between the two species, suggesting that they have different strategies for coping with temperature stress. *Aphaenogaster carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in metabolism, cellular organization and repair. Altogether, these results suggest that *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Temperature plays a prominent role in structuring biological diversity from molecular to macroecological scales [@kingsolver2009]. The ability of individuals to maintain function and survive thermal extremes, thermal tolerance, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. While organisms differ in fixed traits such as proteome stability [@ghosh2010] and behavior [@sunday2014] that allow them to function in the typical range of temperatures to which they are exposed, tolerance to thermal extremes often involves rapid changes in gene expression [@kultz2005; @somero2011]. These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. How this variation of thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], drought tolerance [@villar2011] and many other traits. These RNAseq studies have used an ANOVA-type experimental or sampleing design with only a few levels of temperature or other environmental stressors tested. However, many ecologically-relevant aspects of the environment such as temperature vary continuously in nature. The reaction norm describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014]. Murren et al [-@murren2014] concluded that many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is.

Moreover, the results provided by RNAseq data do not always directly address the aims of studies. For evolutionary ecology studies, the goal of RNAseq studies often is to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014]. However, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005]. Genome-wide association (GWA) studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013]. Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues. Thus, plant and animal breeders have largely abandoned GWA in favor of genomic prediction [@goddard2007; @yang2010].

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which is a characterization of the reaction norm for all genes in an organism's transcriptome across an ecologically-relevant thermal gradient. Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes of two closely-related temperate-zone species to gain insight into how spatially divergent selection has changed patterns of thermally-responsive gene expression. We apply this approach in a comparative study of two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996], which have more southern and northern distributions, respectively, and diverged approximately .... years ago (bernice??). Temperature is a potentially strong selective force on ground-nesting *Aphaenogaster* colonies, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@lubertazzi2012]. We collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina (cite) paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Information).

As *A. carolinensis* lives in a climate that is hotter and has a longer growing season than that of *A. picea* (Supplemental Information), we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways. First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

In addition, responsive genes may change in their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero2011]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes. Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance. Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*, and vice versa.

Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011].

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes, of which 47% had a best hit to an organism in the class Insecta.  Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that were thermally-responsive at 5% False Discovery Rate. Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive. However, the closest related ant species with sequenced genomes have only about 17,00 genes [@gadau2012]. Assuming that the number of genes is similar for *Aphaenogaster*, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.

```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. The total number of genes for two categories (*A. picea* Low - *A. carolinensis* Low; *A. picea* NotResp - *A. carolinensis* Bimodal) is labeled in two boxes for perspective."}
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
```

For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C. Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differs between the two species. While some of these were simply responsive in only one species, `r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes that were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` that were responsive in *A. picea* but not *A. carolinensis*, most of these were genes that changed from one expression category to another between the species (e.g. the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1).

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes. The **High** genes in *A. carolinensis* were activated on average at `r T_on.high.ttest$estimate[1]`°C compared to `r T_on.high.ttest$estimate[2]`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at **Low** temperatures, there was no difference in the mean temperature of gene activation (*t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

@@

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures. Bars are Probability density function of peak expression for all thermally-responsive in *A. carolinensis* (blue) and *A. carolinensis* (red) species"}

@@

As the molecular pathways involved in thermal response to cold and hot temperatures differ, we used gene set enrichment analysis on each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further used GSEA for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A. carolinensis* to having increased expression in **Low** temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For **Intermediate** genes in *A. carolinensis* that shifted **Bimodal** in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

@@

Thermal tolerance is a key trait for determining the survival of individual organisms, the persistence of populations and the geographic limits of species. In this study, we applied a novel regression design to characterize the thermal reactionome of parapatric ant species, *Aphaenogaster picea* and *Aphaenogaster carolinensis*, found in colder and warmer environments, respectively, that are known to differ in their thermal tolerances [@warren2013]. To date, transcriptomic studies have used differential expression analysis between a few different conditions or environments (but see [@hodgins-davis2012]). By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. Specifically, we found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive.

>>>>

---
title: Investigations of thermal reactionomes reveals stress tolerance vs. resistance
  warm and cold-climate ant species
author: \centering{John Stanton-Geddes \textsuperscript{1}, Andrew Nguyen \textsuperscript{1},
  James Vincent \textsuperscript{2}, Mahesh Vangala \textsuperscript{2}, Robert R.
  Dunn \textsuperscript{3}, Nathan J. Sanders \textsuperscript{4}, Nicholas J. Gotelli
  \textsuperscript{1}, Sara Helms Cahan \textsuperscript{1} \\ \vspace{5 mm} \textsuperscript{1}
  Department of Biology, University of Vermont \\ \textsuperscript{2} Vermont Genetics
  Network, University of Vermont \\ \textsuperscript{3} Department of Biological Sciences,
  North Carolina State University \\ \textsuperscript{4} Center for Macroecology,
  Evolution and Climate, University of Copenhagen}
csl: ecology.csl
output:
  word_document: default
  pdf_document:
    fig_caption: yes
    includes:
      in_header: header.tex
    keep_tex: yes
    latex_engine: xelatex
    number_sections: yes
    toc_depth: 2
  html_document: null
includes: header.tex
bibliography: ApTranscriptome.bib
---

@@

Thermal tolerance is a key trait governing the distributions of species and underlies their future responses to climate change. Here, we characterize the thermal “reactionome” – the reaction norm for all genes in an organism's transcriptome measured across a thermal gradient to understand the evolution of thermal tolerance. We characterized the reactionomes of two common ant species in eastern U.S. forests: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. Interestingly, the warm-adapted *A. carolinensis* displayed higher basal levels of gene expression, upregulated fewer heat shock proteins, and exhibited greater inertia of gene expression change to thermal extremes than did *A. picea*. This pattern is consistent with both constitutive thermal defenses and enhanced proteome stability. *A. carolinensis* shut down more genes than *A. picea* in response to thermal extremes, whereas the same genes tended to be upregulated in *A. picea*. These included genes functionally involved in metabolism, cellular organization and repair \todo{Sara - where did the 'repair' term come from?}. These differences between species suggest that the two closely related species employ different strategies for coping with temperature stress, namely *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. More broadly, this study adds to a growing body of evidence that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Temperature regulates biological activity and shapes diversity from molecular to macroecological scales [@brown2004; @kingsolver2009]. Thermal tolerance, the ability of individuals to maintain function and survive thermal extremes, varies considerably among species and populations [@cowles1939; @deutsch2008; @kingsolver2013; @hoffmann2013]. These differences within and among species in thermal tolerance are largely genetic [@krebs1997; @kellermann2012] with a highly polygenic basis [@krebs1998; @williams2012; @morgan2006; @takahashi2011]. How this variation in thermal tolerance evolved, and its potential for future evolution [@hoffmann2008], is not well understood. To address this, we need information on the extent to which geographic differentiation and selection has acted upon the diversity of genes, and their plasticity of expression, involved in thermal tolerance.

In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes underlying thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014], among other traits. These RNAseq studies have used an ANOVA-type experimental or sampling design with only a few levels of temperature. However, temperature varies continuously in nature. As a results, many patterns of biologically-relevant plasticity are likely to be missed with ANOVA-style designs that compare only a few discrete levels of an environmental variable [@murren2014]. Similarly, gene expression studies that look at just a few environments (e.g. hot vs. cold) are likely to miss key differences among species that are due not just to whether it is hot, but rather how hot it is. What is needed is a reaction norm approach, which describes variation in the phenotype of a single genotype in response to a continuous environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values of either species or populations [@murren2014].

Regardless of the experimental design or treatment of temperature as a categorical variable with few levels, an important suite of issues often hampers many RNAseq studies that are designed to identify the genetic basis of environmental adaptation [@meyer2011; @teets2012; @barshis2013; @oneil2014]. Namely, lists of differentially-expressed genes derived from RNAseq results are prone to false positives [@gonzalez2013], depend on the genetic background of the organism [@sarup2011], may miss genes with small expression changes that have detectable phenotypic effects [@yan2002; @padovan2013], identify genes that are correlated, but not causative, of phenotypes [@fay2004] and may not be directly informative about organismal status [@feder2005]. Genome-wide association studies have faced a similar problem: phenotypic effects of a small number of statistically-significant genes are over-estimated while small effects of a large number of other causal genes are missed [@beavis1994; @larsson2013]. Our understanding of the genetic basis of traits, and our ability to predict evolutionary potential, are compromised by these issues.

Here, we present a novel method for RNAseq analysis: the *thermal reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across the temperatures it experiences. Rather than focusing on the identity of individual genes with differential expression, we compare overall patterns in the reactionomes to gain insight into how spatially divergent selection has led to patterns thermally-responsive gene expression in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. Temperature is a potentially strong selective force on ground-nesting ant populations, which must tolerate seasonally freezing winters and hot summers. On shorter time scales, individual workers can experience extreme thermal environments when they leave the thermally buffered ant nest to forage for food [@diamond2012].

As the more southern *A. carolinensis* occurs in climates that are hotter with longer growing seasons than that of the more northerly *A. picea*, we expected patterns of gene expression in response to thermal extremes to differ between these species in a number of ways. First, the number of genes showing plasticity of expression to thermal extremes may differ between the species. Finding that more genes are up-regulated to high temperatures in the colder-climate species *A. picea* would indicate that this species relies on activation of the cellular stress response [@kultz2005] and other protective genes to a greater extent than *A. carolinensis*, while the opposite would be true for cold temperatures. Conversely, finding that more genes are down-regulated in response to thermal extremes would indicate that a species relies to a greater extent upon proteome stability [@ghosh2010] and behavioral mechanisms [@sunday2014] to tolerate thermal stress.

Second, responsive genes may change in their sensitivity to thermal extremes. Finding that heat-responsive genes are up-regulated at higher temperatures in *A. carolinensis* than in *A. picea* would be consistent with local adaptation for greater heat tolerance. That is, more extreme temperatures are required for activation of the heat shock response [@somero2011]. Conversely, finding cold-responsive genes are up-regulated at lower temperatures in *A. picea* than *A. carolinensis* would indicate local adaptation for cold tolerance. Similar patterns may be observed for genes that are down-regulated in response to thermal extremes. Finally, changes in the expression patterns of responsive genes between species can provide insight on the evolution of thermal tolerance. Specifically, if adaptation to thermal extremes works by genetic assimilation [@pigliucci2006], genes up-regulated in response to cold temperatures in *A. carolinensis* may have increased basal expression in the colder-climate species *A. picea*, and vice versa.

Importantly, none of the above hypotheses depend on the functional annotation of the responsive genes, which can be inaccurate in non-model systems [@primmer2013; @skunca2012]. However, insight on the molecular toolkit involved in adaptation to thermal extremes can be gained through circumspect examination of the functional annotation of responsive genes using gene set enrichment analysis. For example, we expect genes in the cellular stress response [@kultz2005] to be up-regulated in response to thermal extremes, whereas we expect genes involved with cell profliferation and non-essential molecular processes to be down-regulated in response to thermal extremes [@somero2011]. To test these predictions for these two exemplar species, we collected RNAseq data at 12 temperatures (every 3.5°C between 0° and 38.5°C) using Illumina paired-end short read data, generated the transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis.

@@

The combined *Aphaenogaster* transcriptome assembly contained 99,861 transcripts, hereafter referred to as genes.  Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` genes with expression patterns that were thermally-responsive at 5% False Discovery Rate. Thus, at a minimum `r round(9100/99861, 2)*100`% of genes are thermally-responsive. However, the closest related ant species with sequenced genomes have only about 17,000 genes [@gadau2012]. Assuming that the number of genes is similar for these two *Aphaenogaster* species, this suggests that up to 50% of the genome (9,100 / 17,000) could be thermally-responsive.

```{r mosaic_plot, echo=FALSE, results='asis', fig.cap = "Mosaic plot of the number of thermally-responsive genes by expression type. **Low** are genes with increased expression at low temperatuers (< 10°C), **Intermediate** are genes with maximum expression between 10 - 30°C, **High** are genes with increased expression at high temperatures (> 31°C), **Bimodal** are genes with increased expression at both low and high temperatures, while **NotResp** are genes that are thermally-responsive in one species but not the other. The width of the columns is proportional to the number of genes for each expression type for *A. carolinensis* while the height of the rows is proportional to the number of genes for each expression type for *A. picea.* The rows are colored by expression type for *A. picea*. **Low** The total number of genes for two categories (*A. picea* **Low** - *A. carolinensis* **Low**; *A. picea* **NotResp** - *A. carolinensis* **Bimodal**) is labeled in two boxes for perspective."}
mosaic.colors <- brewer.pal(5, "Blues")
par(mar = c(5.1, 3.1, 2.1, 7.5))
mosaicplot(tt2, ylab = expression(italic("A. picea")), xlab = "", main = "", color = mosaic.colors, cex.axis = 0.01)
labels <- c("Low", "Intermediate", "High", "Bimodal", "NotResp")
text(c(.1,.5,.75,.85,.99), par("usr")[3] - 0.05, srt = 45, adj = 1, labels = labels, xpd = TRUE, font = 1)
text(.5, par("usr")[4]+.1, "A. carolinensis", xpd = TRUE, font = 3)
text(.15, .7, "2,594")
text(.85, .03, "223", cex = .75)
legend(1.05, .9, fill = mosaic.colors, c("Low", "Intermediate", "High", "Bimodal", "NotResp"), xpd=TRUE)
```

For each species, the thermally-responsive genes were grouped into five expression categories based on their reaction norms: genes that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally-responsive in that species (*NotResp*). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at the mean temperature of 19.5°C. Of the thermally-responsive genes, `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` had an expression pattern that differed between the two species. While some of these were responsive in only one species (`r colSums(tt2)[which(colnames(tt2) == "NotResp")]` genes were responsive in *A. carolinensis* but not *A. picea*, and `r rowSums(tt2)[which(rownames(tt2) == "NotResp")]` were responsive in *A. picea* but not *A. carolinensis*), most of these were genes that changed from one expression category to another between the species (e.g. the number of genes in each expression category differed significantly between species; $\chi^2$ = `r round(chi1$statistic, 0)`, P < `r max(chi1$p.value, 0.001)`; Fig. 1). In both species, the majority of genes were in the **Low** category (Fig. 1). For *A. carolinensis*, this was followed by **Intermediate**, **High** and then **Bimodal**, whereas for *A. picea* the second largest category was **Bimodal** followed by **High** and finally **Intermediate** (Fig. 1).

@@

To examine how the sensitivity of gene expression to changes in temperature differed between the two species, we compared the temperature of gene activation for **High** and **Low** genes. The **High** genes in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~df=`r T_on.high.ttest$parameter`~ = `r T_on.high.ttest$statistic`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For genes with increased expression at **Low** temperatures, there was a negligible difference in the mean temperature of gene activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~df=`r T_on.low.ttest$parameter`~ = `r T_on.low.ttest$statistic`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

@@

```{r T_on_plot, echo=FALSE, warning = FALSE, fig.cap = "Histogram with smooth density estimate of temperature of maximum change in expression for genes that have increased expression at **Low** and **High** temperatures. Red bars and lines are for *A. carolinensis* while blue bars and lines are for *A. picea*."}

@@

To describe the molecular toolkit underlying *Aphaenogaster* response to thermal extremes, we used gene set enrichment analysis to identify gene ontology (GO) categories enriched for each expression category separately. First, we looked at the genes shared in each expression category (e.g. the intersection) for each species (Table S1). Out of a total of `r length(score(Ap.BP.resultParentChild))` GO terms scored, we found `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "High"), ])` enriched (*P* < 0.01) terms for **High** genes. Clustering of terms by semantic similarity yielded 3 clusters, with 5 of the 7 terms being involved in "regulation of biosynthetic/metabolic process". For **Low** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Low"), ])` GO terms were enriched. These terms grouped into 15 clusters, including "protein modification", "protein metabolic process", "lipid metabolic process", "regulation of signaling", "protein transport/localization", "detection of stimulus", and "chondroitin sulfate biosynthesis". For **Bimodal** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Bimodal"), ])` GO terms were enriched. These terms grouped into 5 clusters, including "biological regulation", "regulation of cellular biosynthesis", "gene expression", "signaling", and "protein modification process". For **Intermediate** genes, `r nrow(Ap.gsea.intersect[which(Ap.gsea.intersect$Type == "Intermediate"), ])` GO terms were enriched. These terms grouped into 3 clusters, including "lipid biosynthesis/metabolism", "glycoprotein biosynthesis/metabolism", and "DNA organization". Overall, GSEA indicates that molecular responses to thermal extremes involve large-scale cellular and organizational changes, though the molecular responses to high temperatures are more limited than those to low temperatures.

We further used gene set enrichment analysis for the genes that shifted from being shut-down at extreme temperatures, **Intermediate**, in *A. carolinensis* to having increased expression in **Low** temperatures in *A. picea*. This group included `r nrow(Ar.int.A22.low.gsea)` GO terms that grouped into 12 clusters, including "protein metabolic process", "glycosylation" "transport", "lipid metabolic process", "GTP metabolism", "positive regulation of cellular process" and "organelle organization". For **Intermediate** genes in *A. carolinensis* that shifted **Bimodal** in *A. picea*, there were `r nrow(Ar.int.A22.bim.gsea)` enriched GO terms including "phospholipid catabolic process", "L-amino acid transport", "macromolecule metabolic process", "gluatmate receptor signaling pathway" and "histone modification". These results indicate that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, shuts down metabolism to a greater extent than the colder-climate species, *A. picea*, which increases metabolism.

@@

By using a regression-approach, we are able to quantitatively describe changes in gene expression, rather than simply report the numbers of differentially-expressed genes. Specifically, we found that at least 10%, and potentially 50%, of the genes in these species were thermally-responsive. This result is consistent with the only other work we are aware of that examines reaction norms of gene expression, which found that most genes in yeast have variable expression in response to copper exposure [@hodgins-davis2012]. In general, finding thousands of genes that respond to temperature highlights the limitations of previous methods that only found a few dozen to hundred genes are involved in thermal response from recent studies in corals [@meyer2011; @barshis2013; @palumbi2014] and butterflies [@oneil2014]. Underlying the importance of thermal tolerance as a physiological trait [@kingsolver2009], the genetic basis of thermal tolerance is likely to be much more complex than indicated by these previous studies.

The patterns of gene expression between species differed substantially, with more genes down-regulated in *A. carolinensis* and more up-regulated in *A. picea* in response to thermal extremes (Fig. 1, 3). The results of the gene set enrichment analysis give an indication of what might drive this difference. The genes that are shut-down by *A. carolinensis* in response to cold temperatures, but up-regulated by *A. picea* include gene ontology clusters involved in metabolism, transport and regulation. Thus, *A. carolinensis*, the more southerly species, shuts-down metabolism to a lesser extent than *A. picea* and so must rely on other mechanisms such as behavior (e.g. retreat into its nest) or proteome stability to tolerate thermal stress. In contrast, *A. picea* increases metabolism to a greater extent than *A. carolinensis* with more extreme temperatures, potentially to mount a greater molecular response to thermal stress. Consistent with this hypothesis, we found that of the 7 thermally-responsive heat shock proteins, all 7 increase expression in response to thermal extremes in *A. picea*, but only 4 do in *A. carolinensis*. This result supports our expectation that *A. picea* would rely to a greater extent upon activation of the cellular stress response and other protective genes to a greater extent than *A. carolinensis*. Our results are at least congruent with evidence from other systems [@barshis2013] that thermally-stressful habitats select for investment in tolerance, while organisms from less stressful environments rely on plastically-induced resistance. ...Sara - something about enzyme efficiency tradeoffs???
